Picky with peakpicking: assessing chromatographic peak quality with simple

### 2023-06-06 manuscript/manuscript.Rmd
Switched from GPF to GFF and recompiled everything

The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GPF values implying that a large majority of the good MFs passed the threshold with very little noise included. The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GPF values closer to 75%. The two-parameter model performed worst when tested internally, with an FDR of about 20% and %GPF also around 75% (Figure 2). However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and %GPF around 60-80. The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more MFs recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).

```{r Fig2 FDR vs GPF cross-train triangle figure feature selection}
fdr_gpf_cross_train_df <- lst(all_params, raw_data_params, xcms_params) %>%
  imap(function(param_selection, param_name){
    lapply(dataset_versions, function(train_set){
      full_model <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        select(feat_class, all_of(param_selection)) %>%
        mutate(feat_class=feat_class=="Good") %>%
        glm(formula=feat_class~., family = binomial)
      lapply(dataset_versions, function(test_set){
        get(test_set) %>%
          mutate(pred_prob=predict(object=full_model,newdata = ., type = "response")) %>%
          mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
          select(feat_class, starts_with("pred_class")) %>%
          mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
          mutate(train_test=str_remove_all(train_test, "_features")) %>%
          mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
      }) %>% bind_rows()
    }) %>% bind_rows() %>% mutate(which_params=param_name)
  }) %>% bind_rows() %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  group_by(train_test, which_params, cross_type) %>%
  count(error_type) %>%
  ungroup() %>%
  mutate(cross_type=factor(cross_type, levels=c("Same", "Different"))) %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(FDR=FP/(FP+TP)) %>%
  mutate(GPF=TP/(FN+TP))

train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
fdr_gpf_cross_train_gp <- fdr_gpf_cross_train_df %>%
  pivot_longer(c(FDR, GPF)) %>%
  mutate(name=paste0("%", name)) %>%
  mutate(which_params=factor(which_params, 
                             labels=c("All metrics", "XCMS metrics", "Raw data metrics"),
                             levels=c("all_params", "xcms_params", "raw_data_params"))) %>%
  mutate(cross_type_num=as.numeric(cross_type)) %>%
  mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
  mutate(name=factor(name, levels=c("%GPF", "%FDR"), labels=c("GPF", "FDR"))) %>%
  ggplot() +
  geom_point(aes(x=cross_type, y=value, color=train_test, 
                 fill=train_test, shape=name), size=5) +
  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
              method = "lm", formula="y~x", color="black", se=FALSE) +
  facet_grid(~which_params) +
  scale_x_discrete(labels = c("Same\ndataset\n", "Different\ndataset\n"),
                   breaks = c("Same", "Different")) +
  scale_color_manual(values = c("#0b505c", "#184d1f", "#0d8299", "#028e34"), 
                     breaks = train_test_labels,
                     aesthetics = c("color", "fill")) +
  scale_shape_manual(breaks = c("GPF", "FDR"), values = c("\u25B2", "\u25BC")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme() +
  labs(x=NULL, y=NULL, color="Train-test", fill="Train-test", shape="Error type") +
  guides(shape=guide_legend(order=1))

ggsave(plot = fdr_gpf_cross_train_gp, 
       filename = "fdr_gpf_cross_train.png", 
       device = "png", path = "manuscript/figures",
       width = 6.5, height = 3, units = "in", dpi = 300)
```

![Figure 2: False discovery rate (%FDR) and fraction of good peaks found (GPF) plotted across different subsets of model parameters. Lower %FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher %GPF indicates a larger fraction of the total good features were found using the same threshold. Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel. Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gpf_cross_train.png){width=90%}

@@

None of the penalized regression models significantly improved cross-validated performance between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets. All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GPF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp. figure 3). The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GPF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp. figure 3). In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.

@@

With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively (Figure 4). Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good MFs making it past this threshold. In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively. Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GPF but expect it to be in a similar range (Figure 4).

```{r FDR vs GPF with additional datasets}
# Currently a fairly simple figure, may need additional data?
both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
pred_probs_all <- bind_rows(lst(
  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
  CultureData=CultureData_features, Pttime=Pttime_features), 
  .id = "dataset") %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
  # count(pred_class, dataset)
  mutate(error_type=case_when(
    feat_class=="Good" & pred_class=="Good" ~ "%TP",
    feat_class=="Good" & pred_class=="Bad" ~ "%FN",
    feat_class=="Bad" & pred_class=="Good" ~ "%FP",
    feat_class=="Bad" & pred_class=="Bad" ~ "%TN",
    TRUE ~ "Incomparable"
  )) %>%
  # count(feat_class, pred_class, dataset) %>% filter(dataset=="Pttime")
  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
  filter(error_type!="Incomparable") %>%
  group_by(dataset, error_type) %>%
  summarise(n=n(), .groups = 'drop') %>%
  complete(dataset, error_type, fill=list(n=0))
error_type_all_datasets <- pred_probs_all %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(`FDR`=`%FP`/(`%FP`+`%TP`)) %>%
  mutate(`GPF`=`%TP`/(`%FN`+`%TP`)) %>%
  mutate(`GPF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GPF`))

fdr_gpf_all_datasets_gp <- error_type_all_datasets %>%
  pivot_longer(c(`FDR`, `GPF`)) %>%
  drop_na() %>%
  mutate(desc=case_when(
    is.na(value) ~ NA,
    name=="FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
    name=="GPF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
  )) %>%
  mutate(vjust_custom=case_when(
    is.na(value) ~ NA,
    name=="FDR" & dataset=="Pttime" ~ 1.3,
    name=="FDR" ~ -0.3,
    name=="GPF" ~ 1.3,
  )) %>%
  mutate(name=factor(name, levels=c("GPF", "FDR"),
                     labels=c("Fraction of good peaks found (GPF)", 
                              "False discovery rate (FDR)"))) %>%
  mutate(dataset=factor(dataset, 
                        levels = c("Falkor", "MESOSCOPE", "CultureData", "Pttime"),
                        labels= c("Falkor", "MESO-\nSCOPE", "Culture\nData", "Pttime"))) %>%
  ggplot(aes(x=dataset, y=value)) +
  geom_point(size=3) +
  geom_text(aes(label=desc, vjust=vjust_custom), parse=TRUE) +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
  theme_bw() +
  theme(axis.title = element_blank(),
        strip.background = element_rect(fill="white"))

ggsave(plot = fdr_gpf_all_datasets_gp, 
       filename = "fdr_gpf_all_datasets.png", 
       device = "png", path = "manuscript/figures",
       width = 3, height = 4, units = "in", dpi = 300)

@@

![Figure 4: False discovery rate and proportion of total good features identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9. FDR is calculated by dividing the number of false positives by the total positives produced by the model and GPF is calculated by dividing the number of true positives by the total number of good features as identified manually. Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gpf_all_datasets.png){width=30%}

@@

source("manuscript/IS_integrations/is_combine.R")
IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
  mutate(filename=paste0(filename, ".mzML")) %>%
  dplyr::rename(area="IS_area") %>%
  filter(!str_detect(filename, "QC")) %>%
  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
pooled_IS_areas <- IS_areas %>%
  filter(str_detect(filename, "_Poo_"))

@@

clockplot_complete_df <- clockplot_data %>%
  distinct(feature, feat_class, pred_class) %>%
  left_join(clockplot_xy_spots, by = join_by(feature)) %>%
  left_join(kw_pvals_all %>% select(feature, fdr_pval_byclass), by = join_by(feature)) %>%
  left_join(kw_classes_all, by = join_by(feature)) %>%
  mutate(error_type=case_when(
    pred_class=="Good" & feat_class=="Good" ~ "TP",
    pred_class=="Good" & feat_class=="Bad" ~ "FP",
    pred_class=="Bad" & feat_class=="Good" ~ "FN",
    pred_class=="Bad" & feat_class=="Bad" ~ "TN"
  )) %>%
  mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
                           labels=c("Good peak found", "Bad peak included",
                                    "Good peak missed", "Bad peak avoided")))

@@

clockplot_complete_gp <- clockplot_base_gp +
  geom_point(aes(shape=error_type), alpha=0.5, size=2,
             data = clockplot_complete_df) +
  scale_shape_manual(breaks = c("Good peak found", "Bad peak included",
                                "Good peak missed", "Bad peak avoided"), 
                     values = c(19, 3, 4, 1))

@@

We used two fully-labeled and two partially-labeled HILIC LC-MS datasets to assess the performance of the XCMS algorithm and construct a robust model of peak quality. To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good peaks found (GPF, also known as recall or sensitivity) and the percentage of bad molecular features (MFs) included, also known as the false discovery rate (FDR). We decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.

@@

This model outperforms the previously reported logistic regression model in @Kantz2019 and is highly simplified. There, they used a nineteen-parameter multiple logistic regression model and found a maximum performance of 80% GPF and an FDR of 34% on a cross-validated second cohort, similar to our cross-dataset testing. Our final two-parameter model also had an 80% GPF at a 0.5 likelihood threshold, but a significantly lower FDR of ~22.9%. This increased performance is likely due to our use of the metrics recalculated from the raw data, as their metrics were only calculated from the default MzMine2 peak parameters reported: similar to our XCMS-only model.

@@

The class imbalance, with mostly poor quality MFs, is partially why we chose to measure precision and recall instead of total accuracy. However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GPF metrics as well as providing the raw confusion matrices in Supplemental Table XX. @Melnikov2020 reported precision and recall in their presentation of `peakonly`, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GPF and FDR (89% and 3%, respectively, relative to our 77.1% GPF and 19.6% FDR overall). However, if we report precision and recall with the positive class set to "Bad", essentially trying to predict poor-quality MFs instead of good ones, our precision becomes 96.5% and our recall 95.7% due to the strong prior information about most MFs being bad.

@@

maybe_skews <- c(2.5, 3, 4, 5)
skew_names <- c("Large skew", "Small skew", "Minimal skew", "Zero skew")
names(maybe_skews) <- skew_names
perf_peak_data <- single_EIC %>%
  group_by(feature) %>%
  select(feature, rt, int) %>%
  mutate(rt=(rt-min(rt))/(max(rt)-min(rt))) %>%
  mutate(int=(int-min(int))/(max(int)-min(int))) %>%
  cbind(sapply(maybe_skews, function(x){
    perf_peak <- dbeta(.$rt, shape1=x, shape2=5)
    perf_peak/max(perf_peak)
  })) %>%
  dplyr::rename(`Raw data`="int") %>%
  mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
                        labels=c("Good peak", "Bad peak")))
summary_stats <- perf_peak_data %>%
  pivot_longer(-c(feature, rt, `Raw data`)) %>%
  mutate(new_val=value-`Raw data`) %>%
  group_by(feature, name) %>%
  summarise(single_cor=round(cor(`Raw data`, value), 2), 
            single_sd=round(sd(new_val), 2), 
            .groups = "drop_last") %>%
  mutate(name=factor(name, levels=skew_names)) %>%
  mutate(x_coord=c(0.125+(as.numeric(name)-1)/4))

@@

In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good. This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good peaks found (GPF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of actual positives).

@@

GPF: Good Peaks Found

>>>>

The full model performed very well when tested internally on the same dataset both during 80/20 cross validation and when using the full dataset, with FDR values in the 5-10% range and 80-90% GFF values implying that a large majority of the good MFs passed the threshold with very little noise included. The XCMS metrics performed slightly worse, with FDR values in the 10-15% range and GFF values closer to 75%. The two-parameter model performed worst when tested internally, with an FDR of about 20% and GFF also around 75% (Figure 2). However, when the models were trained on a different dataset than the one they were used to predict classifications for, all models had similar performance with %FDR around 10-25 and GFF around 60-80. The model trained on MESOSCOPE and tested on Falkor had consistently higher values, indicating that it was favoring more MFs recovered at the cost of a higher FDR, while the reverse was true for the model trained on Falkor and tested on MESOSCOPE (Figure 2).

```{r Fig2 FDR vs GFF cross-train triangle figure feature selection}
fdr_gff_cross_train_df <- lst(all_params, raw_data_params, xcms_params) %>%
  imap(function(param_selection, param_name){
    lapply(dataset_versions, function(train_set){
      full_model <- get(train_set) %>% 
        filter(feat_class%in%c("Good", "Bad")) %>%
        select(feat_class, all_of(param_selection)) %>%
        mutate(feat_class=feat_class=="Good") %>%
        glm(formula=feat_class~., family = binomial)
      lapply(dataset_versions, function(test_set){
        get(test_set) %>%
          mutate(pred_prob=predict(object=full_model,newdata = ., type = "response")) %>%
          mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
          select(feat_class, starts_with("pred_class")) %>%
          mutate(train_test=paste(train_set, test_set, sep = "-")) %>%
          mutate(train_test=str_remove_all(train_test, "_features")) %>%
          mutate(cross_type=ifelse(train_set==test_set, "Same", "Different"))
      }) %>% bind_rows()
    }) %>% bind_rows() %>% mutate(which_params=param_name)
  }) %>% bind_rows() %>%
  mutate(error_type=case_when(
    feat_class=="Bad" & pred_class=="Bad" ~ "TN",
    feat_class=="Bad" & pred_class=="Good" ~ "FP",
    feat_class=="Good" & pred_class=="Bad" ~ "FN",
    feat_class=="Good" & pred_class=="Good" ~ "TP"
  )) %>%
  group_by(train_test, which_params, cross_type) %>%
  count(error_type) %>%
  ungroup() %>%
  mutate(cross_type=factor(cross_type, levels=c("Same", "Different"))) %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(FDR=FP/(FP+TP)) %>%
  mutate(GFF=TP/(FN+TP))

train_test_levels <- c("MS3000-MS3000", "FT2040-FT2040", "MS3000-FT2040", "FT2040-MS3000")
train_test_labels <- c("MESO-MESO", "Falkor-Falkor", "MESO-Falkor", "Falkor-MESO")
fdr_gff_cross_train_gp <- fdr_gff_cross_train_df %>%
  pivot_longer(c(FDR, GFF)) %>%
  mutate(name=paste0("%", name)) %>%
  mutate(which_params=factor(which_params, 
                             labels=c("All metrics", "XCMS metrics", "Raw data metrics"),
                             levels=c("all_params", "xcms_params", "raw_data_params"))) %>%
  mutate(cross_type_num=as.numeric(cross_type)) %>%
  mutate(train_test=factor(train_test, levels=train_test_levels, labels=train_test_labels)) %>%
  mutate(name=factor(name, levels=c("%GFF", "%FDR"), labels=c("GFF", "FDR"))) %>%
  ggplot() +
  geom_point(aes(x=cross_type, y=value, color=train_test, 
                 fill=train_test, shape=name), size=5) +
  geom_smooth(aes(x=cross_type_num, y=value, group=name), 
              method = "lm", formula="y~x", color="black", se=FALSE) +
  facet_grid(~which_params) +
  scale_x_discrete(labels = c("Same\ndataset\n", "Different\ndataset\n"),
                   breaks = c("Same", "Different")) +
  scale_color_manual(values = c("#0b505c", "#184d1f", "#0d8299", "#028e34"), 
                     breaks = train_test_labels,
                     aesthetics = c("color", "fill")) +
  scale_shape_manual(breaks = c("GFF", "FDR"), values = c("\u25B2", "\u25BC")) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(labels = scales::label_percent()) +
  theme_bw() +
  theme() +
  labs(x=NULL, y=NULL, color="Train-test", fill="Train-test", shape="Error type") +
  guides(shape=guide_legend(order=1))

ggsave(plot = fdr_gff_cross_train_gp, 
       filename = "fdr_gff_cross_train.png", 
       device = "png", path = "manuscript/figures",
       width = 6.5, height = 3, units = "in", dpi = 300)
```

![Figure 2: False discovery rate (%FDR) and fraction of good features found (%GFF) plotted across different subsets of model parameters. Lower %FDR indicates a smaller fraction of false positives among those molecular features the model categorized as "Good" using a threshold of 0.5, and higher %GFF indicates a larger fraction of the total good features were found using the same threshold. Points are colored by the model used for training and testing, with internal validation (using the same dataset for training as prediction) in the darker colors on the left and external validation (using a different dataset for training than prediction) in the lighter colors on the right of each panel. Lines of best fit have been estimated and plotted in black on top of the data points.](figures/fdr_gff_cross_train.png){width=90%}

@@

None of the penalized regression models significantly improved cross-validated performance between the MESOSCOPE and Falkor datasets when measured by both initial performance and the performance drop when applied across datasets. All three regularized regression models had similar behavior, with ridge regression (α = 0) had the lowest rates for both GFF and FDR, with lasso (α = 1) obtaining higher ones and representing a less-stringent false negative acceptance while, as expected, the elastic net (α = 0.5) fell in between the two (Supp. figure 3). The random forest model, interestingly, had perfect predictive capacity when tested internally on the training data (FDR=0%, GFF=100%, for both MESOSCOPE and Falkor) but showed a significant drop in improvement when applied across datasets (Supp. figure 3). In each case, the performance drop when applied to a novel dataset was more extreme than the simple two-parameter model described above.

@@

With the stricter threshold, we obtained FDR values consistently below 5% even on the novel datasets, with values of 1.0%, 0.0% (truly zero false positives), 2.0%, and 4.6% for Falkor, MESOSCOPE, CultureData, and Pttime respectively (Figure 4). Of course, this low error rate required that we miss out on additional potentially valuable features, with only a fraction of the total good MFs making it past this threshold. In both the Falkor and MESOSCOPE datasets, less than half of the good MFs were labeled as such, with actual values of 39.4% and 26.5%, respectively. Since we did not label the complete dataset for CultureData and Pttime, we cannot accurately calculate the %GFF but expect it to be in a similar range (Figure 4).

```{r FDR vs GFF with additional datasets}
# Currently a fairly simple figure, may need additional data?
both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)
pred_probs_all <- bind_rows(lst(
  Falkor=FT2040_features, MESOSCOPE=MS3000_features,
  CultureData=CultureData_features, Pttime=Pttime_features), 
  .id = "dataset") %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.9, "Good", "Bad")) %>%
  # count(pred_class, dataset)
  mutate(error_type=case_when(
    feat_class=="Good" & pred_class=="Good" ~ "%TP",
    feat_class=="Good" & pred_class=="Bad" ~ "%FN",
    feat_class=="Bad" & pred_class=="Good" ~ "%FP",
    feat_class=="Bad" & pred_class=="Bad" ~ "%TN",
    TRUE ~ "Incomparable"
  )) %>%
  # count(feat_class, pred_class, dataset) %>% filter(dataset=="Pttime")
  mutate(dataset=factor(dataset, levels=c("MESOSCOPE", "Falkor", "CultureData", "Pttime"))) %>%
  filter(error_type!="Incomparable") %>%
  group_by(dataset, error_type) %>%
  summarise(n=n(), .groups = 'drop') %>%
  complete(dataset, error_type, fill=list(n=0))
error_type_all_datasets <- pred_probs_all %>%
  pivot_wider(names_from=error_type, values_from = n) %>%
  mutate(`FDR`=`%FP`/(`%FP`+`%TP`)) %>%
  mutate(`GFF`=`%TP`/(`%FN`+`%TP`)) %>%
  mutate(`GFF`=ifelse(dataset%in%c("CultureData", "Pttime"), NA, `GFF`))

fdr_gff_all_datasets_gp <- error_type_all_datasets %>%
  pivot_longer(c(`FDR`, `GFF`)) %>%
  drop_na() %>%
  mutate(desc=case_when(
    is.na(value) ~ NA,
    name=="FDR"~ paste0("frac(",`%FP`,",",`%FP`+`%TP`,")"),
    name=="GFF"~ paste0("frac(",`%TP`,",",`%FN`+`%TP`,")")
  )) %>%
  mutate(vjust_custom=case_when(
    is.na(value) ~ NA,
    name=="FDR" & dataset=="Pttime" ~ 1.3,
    name=="FDR" ~ -0.3,
    name=="GFF" ~ 1.3,
  )) %>%
  mutate(name=factor(name, levels=c("GFF", "FDR"),
                     labels=c("Fraction of good features found (GFF)", 
                              "False discovery rate (FDR)"))) %>%
  mutate(dataset=factor(dataset, 
                        levels = c("Falkor", "MESOSCOPE", "CultureData", "Pttime"),
                        labels= c("Falkor", "MESO-\nSCOPE", "Culture\nData", "Pttime"))) %>%
  ggplot(aes(x=dataset, y=value)) +
  geom_point(size=3) +
  geom_text(aes(label=desc, vjust=vjust_custom), parse=TRUE) +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
  theme_bw() +
  theme(axis.title = element_blank(),
        strip.background = element_rect(fill="white"))

ggsave(plot = fdr_gff_all_datasets_gp, 
       filename = "fdr_gff_all_datasets.png", 
       device = "png", path = "manuscript/figures",
       width = 3, height = 4, units = "in", dpi = 300)

@@

![Figure 4: False discovery rate and proportion of total good features identified as good by the two-parameter model trained on the combined MESOSCOPE/Falkor dataset and applied to each dataset individually with a threshold of 0.9. FDR is calculated by dividing the number of false positives by the total positives produced by the model and GFF is calculated by dividing the number of true positives by the total number of good features as identified manually. Points correspond to the calculated percentage and absolute numbers are provided above/below the point.](figures/fdr_gff_all_datasets.png){width=30%}

@@

source("manuscript/IS_integrations/is_combine.R")
IS_areas <- read_csv("manuscript/IS_integrations/all_IS.csv") %>%
  mutate(filename=paste0(filename, ".mzML")) %>%
  dplyr::rename(IS_area="area") %>%
  filter(!str_detect(filename, "QC")) %>%
  pivot_wider(names_from=IS_name, values_from = "IS_area") %>%
  mutate("Inj_vol"=ifelse(str_detect(filename, "Poo.*Half"), 0.5, 1)) %>%
  pivot_longer(-filename, names_to = "IS_name", values_to = "IS_area")
pooled_IS_areas <- IS_areas %>%
  filter(str_detect(filename, "_Poo_"))

@@

clockplot_complete_df <- clockplot_data %>%
  distinct(feature, feat_class, pred_class) %>%
  left_join(clockplot_xy_spots, by = join_by(feature)) %>%
  left_join(kw_pvals_all %>% select(feature, fdr_pval_byclass), by = join_by(feature)) %>%
  left_join(kw_classes_all, by = join_by(feature)) %>%
  mutate(error_type=case_when(
    pred_class=="Good" & feat_class=="Good" ~ "TP",
    pred_class=="Good" & feat_class=="Bad" ~ "FP",
    pred_class=="Bad" & feat_class=="Good" ~ "FN",
    pred_class=="Bad" & feat_class=="Bad" ~ "TN"
  )) %>%
  mutate(error_type=factor(error_type, levels=c("TP", "FP", "FN", "TN"),
                           labels=c("Good MF found", "Bad MF included",
                                    "Good MF missed", "Bad MF avoided")))

@@

clockplot_complete_gp <- clockplot_base_gp +
  geom_point(aes(shape=error_type), alpha=0.5, size=2,
             data = clockplot_complete_df) +
  scale_shape_manual(breaks = c("Good MF found", "Bad MF included",
                                "Good MF missed", "Bad MF avoided"), 
                     values = c(19, 3, 4, 1))

@@

We used two fully-labeled and two partially-labeled HILIC LC-MS datasets to assess the performance of the XCMS algorithm and construct a robust model of peak quality. To measure performance, we used two measurements of success closely tied to intuitive questions about a dataset: the percentage of total good features found (GFF, also known as recall or sensitivity) and the percentage of bad molecular features (MFs) included, also known as the false discovery rate (FDR). We decided against using the F$_1$ score as an overall summary statistic because false negatives and false positives have very different implications in this context and should be treated separately.

@@

This model outperforms the previously reported logistic regression model in @Kantz2019 and is highly simplified. There, they used a nineteen-parameter multiple logistic regression model and found a maximum performance of 80% GFF and an FDR of 34% on a cross-validated second cohort, similar to our cross-dataset testing. Our final two-parameter model also had an 80% GFF at a 0.5 likelihood threshold, but a significantly lower FDR of ~22.9%. This increased performance is likely due to our use of the metrics recalculated from the raw data, as their metrics were only calculated from the default MzMine2 peak parameters reported: similar to our XCMS-only model.

@@

The class imbalance, with mostly poor quality MFs, is partially why we chose to measure precision and recall instead of total accuracy. However, precision and recall can also be ambiguous when the positive class is not specified and the raw confusion matrices are unavailable, thus our very precise use of the FDR and GFF metrics as well as providing the raw confusion matrices in Supplemental Table XX. @Melnikov2020 reported precision and recall in their presentation of `peakonly`, relative to which we obtained higher accuracy (they report 89% accuracy) but worse GFF and FDR (89% and 3%, respectively, relative to our 77.1% GFF and 19.6% FDR overall). However, if we report precision and recall with the positive class set to "Bad", essentially trying to predict poor-quality MFs instead of good ones, our precision becomes 96.5% and our recall 95.7% due to the strong prior information about most MFs being bad.

@@

maybe_skews <- c(2.5, 3, 4, 5)
skew_names <- c("Large skew", "Small skew", "Minimal skew", "Zero skew")
names(maybe_skews) <- skew_names
perf_peak_data <- single_EIC %>%
  group_by(feature) %>%
  select(feature, rt, int) %>%
  mutate(rt=(rt-min(rt))/(max(rt)-min(rt))) %>%
  mutate(int=(int-min(int))/(max(int)-min(int))) %>%
  cbind(sapply(maybe_skews, function(x){
    perf_peak <- dbeta(.$rt, shape1=x, shape2=5)
    perf_peak/max(perf_peak)
  })) %>%
  dplyr::rename(`Raw data`="int") %>%
  mutate(feature=factor(feature, levels=c("FT0588", "FT0280"), 
                        labels=c("Good MF", "Bad MF")))
summary_stats <- perf_peak_data %>%
  pivot_longer(-c(feature, rt, `Raw data`)) %>%
  mutate(new_val=value-`Raw data`) %>%
  group_by(feature, name) %>%
  summarise(single_cor=round(cor(`Raw data`, value), 2), 
            single_sd=round(sd(new_val), 2), 
            .groups = "drop_last") %>%
  mutate(name=factor(name, levels=skew_names)) %>%
  mutate(x_coord=c(0.125+(as.numeric(name)-1)/4))

@@

In each case, we categorized each molecular feature as a true positive (TP) if it was predicted to be Good and was manually classified as Good, a true negative if both predicted and classifed as Bad, a false positive if predicted to be Good but manually classified as Bad, and a false negative if predicted to be Bad but was in fact manually classified as Good. This allowed us to additionally define two useful measures of success, the traditionally-defined false discovery rate (FDR, defined as 1-precision or the number of false positives divided by the total number of predicted positives) and the percentage of good features found (GFF, also known as the recall or sensitivity and defined as the number of true positives divided by the total number of actual positives).

@@

GFF: Good Feature Found
